Effects of Trauma Life Support Training Programs on Patient Outcomes: A Pilot Cluster Randomised Trial

### 2024-09-28 pilot-publication/manuscript.Rmd
Respond to all comments

The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) and the low-cost alternative Primary Trauma Care (PTC) are the most established trauma life support training programmes with physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018].Observational studies indicate that these programmes may improve patient outcomes [@Jin2021], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020].

Several systematic reviews call for trials in settings where these programmes are not routinely taught, but data needed to inform the planning of such a trial is missing. We performed a pilot study to assess the feasibility of conducting a cluster randomised controlled trial comparing ATLS^®^ and PTC with standard care, and to estimate probable effect sizes and the intracluster correlation coefficient needed for the sample size calculations of a full-scale trial.

@@

## Variables
The research officers collected data on demographics, time of injury to arrival at the participating hospital, time to recording vital signs, vital signs, times to and management details including imaging and surgery, and details of any injury sustained. <!--Supplemental Material 2 contains a full variable list, with definitions.-->

@@

We found that the ATLS^®^ arm had lower 30-day mortality compared to the PTC and standard care arms. We also found that the PTC arm had lower mortality than the standard care arm. These findings indicate a large potential effect of training physicians in trauma life support, but it is important to note that this pilot study was not powered to detect any differences in outcomes. The arms differed considerably in sample size, with the ATLS^®^ arm having the smallest sample size. This difference most likely resulted from the randomisation process with a small number of heterogeneous clusters.

@@

Several controlled trials, including at least two randomized
controlled trials [@Ali1996; @Ali1999], show that ATLS^®^ is associated with improved provider skills [@Mohammad2013]. Observational evidence indicates that PTC also leads to improved provider skills [@Kadhum2020]. The missing link is then how, and if, these improved skills translate into improved patient outcomes. As trauma care providers we assume, and probably rightly so, that we deliver better
care if we train. The question is then how we should train, especially considering the costs associated with some of the programs offered.

We conclude that a full-scale cluster randomised trial should be feasible after incorporating the lessons of this pilot, and that this full-scale trial should, regardless of its outcome, influence how we train trauma care providers in the future.

@@

## Trial Registration
We intended to also register our trial with Clinical Trials Registry - India but because of time constrains we had to initiate the study before registration was finalised, and Clinical Trials Registry - India only allow prospective registration.

@@

## Follow up of residents
We stated that resident participants would be followed up 30 days after training, if they are in the intervention arms, or 30 days after the study started, if they are in the control arm, but the intervention period was three months.

## Data collection from records
We decided to record data from records only for a subset of patients to reduce the research officers' workload.

@@

## Timing of resident consent
We had initially planned to ask residents for consent before
randomisation, but the units were only finalised after the hospitals had been randomised, and residents were therefore approached for consent afterwards.

>>>>

The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) and the low-cost alternative Primary Trauma Care (PTC) are the most established trauma life support training programmes with physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018]. Observational studies indicate that these programmes may improve patient outcomes [@Jin2021], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020].

Several studies, including at least two randomised controlled trials [@Ali1996; @Ali1999], show that ATLS^®^ is associated with improved provider skills [@Mohammad2013]. Observational evidence indicates that PTC also leads to improved provider skills [@Kadhum2020]. The missing link is then how, and if, these improved skills translate into improved patient outcomes.

Systematic reviews call for trials in settings where these programmes are not routinely implemented, but data needed to inform the planning of such a trial is missing. We aimed to assess the feasibility of conducting a cluster randomised controlled trial comparing ATLS^®^ and PTC with standard care, and to estimate probable effect sizes and the intracluster correlation coefficient needed for the sample size calculations of a full-scale trial.

@@

## Variables
The research officers collected data on demographics, time of injury to arrival at the participating hospital, time to recording vital signs, vital signs, times to and management details including imaging and surgery, and details of any injury sustained. For a subset of patients we also extracted data from medical records, to be able to compare their distributions with directly observed data.<!--Supplemental Material 2 contains a full variable list, with definitions.-->

@@

We found that the ATLS^®^ and PTC arms had lower 30-day mortality compared to the PTC and standard care arms. These findings hints towards a potential effect of training physicians in trauma life support, but it is important to note that this pilot study was not powered to detect any differences in outcomes. The arms differed considerably in sample size, with the ATLS^®^ arm having the smallest sample size. This difference most likely resulted from the randomisation process with a small number of heterogeneous clusters. This heterogeneity highlights the importance of taken different cluster sizes into account in the design of the full scale trial.

@@

We conclude that a full-scale cluster randomised trial should be feasible after incorporating the lessons of this pilot.

@@

## Trial Registration
We intended to also register our trial with Clinical Trials Registry - India and will do so with the full-scale trial.

@@

## Follow up of residents
We stated that resident participants would be followed up 30 days after training, but revised this to follow them up after the end of the study period.

## Data collection from records
We decided to extract data from medical records only for a subset of patients to reduce the research officers' workload.

@@

## Timing of resident consent
We had initially planned to ask residents for consent before
randomisation, but this was not possible because of logistical issues the units were only finalised after the hospitals had been randomised. Residents were therefore approached for consent after randomisation but before training.
